% !TeX root = main.tex

\centerline{\sectionfont\the\sectcount. Rigidity Results}
\medskip
In this section, we present volume rigidity and diameter rigidity for closed and open manifolds respectively.
\medskip
\noindent{\bf\the\sectcount.\the\subsectcount.\ Volume Rigidity.}
Volume rigidity builds on the following rigidity result for volume comparison inequality.
\medskip
\definexref{prop:volume rigidity}{\propnumber}{prop}
\proclaim Proposition~\propnumber. Let $(M,g)$ be a Riemannian manifold with $\Ric\geq(n-1)H$.
If there is an $R>0$ such that $\vol(B_p(R))=\vol(B^H(R))$, then $B_p(R)$ must be isometric to $B^H(R)$.\par
\noindent{\it Proof.}\/ By volume comparison theorem, the function
$$r\mapsto{\vol(B_p(r))\over\vol(B^H(r))}\leq 1$$
is nonincreasing, so if $\vol(B_p(R))=\vol(B^H(R))$ holds for some $R>0$, then $\vol(B_p(\rho))=\vol(B^H(\rho))$ holds for all $0<\rho<R$.
Clearly we are before the cut locus.
Let's trace back when the equality holds in volume comparison inequality.
We used Cauchy--Schwarz inequality in
$$|\Hess r|^2=|\II|^2\geq{m^2\over n-1},$$
hence if equality holds, $\II$ has equal eigenvalues.
Thus $\partial B_p(\rho)$ is totally umbilical, in particular, $m(\rho,\theta)$ is constant on $\partial B_p(\rho)$.
Under geodesic polar coordinate, with respective to the base point $p$, the metric $g$ has form
$$g=dr^2+h(r,\theta).$$
Since $\partial B_p(\rho))$ is totally umbilical, we have
$$\II=\lambda(r)h$$
for all $0<r<R$.
Notice that
$$\II_{ij}=\langle\nabla_i\partial_r,\partial_j\rangle=\partial_r\langle\partial_i,\partial_j\rangle-\langle\partial_i,\nabla_j\partial_r\rangle=\partial_rh_{ij}-\II_{ij},$$
this implies $\II_{ij}=(1/2)\partial_rh_{ij}$.
Thus we have an ordinary differential equation
$${1\over2}\partial_rh_{ij}=\lambda(r)h_{ij}.$$
Solving this equation, we obtain
$$h_{ij}=(\varphi(r))^2h_{ij}(r_0,\theta).\eqdef{eq:h_ij}$$
We know the asymptotic expansion of Riemannian metric under geodesic polar coordinate is
$$h_{ij}=r^2\sigma_{ij}+O(r^4),$$
where $\sigma$ is the canonical metric of round sphere.
Compare~\eqref{eq:h_ij}, we obtain $h_{ij}=f(r)\sigma_{ij}$, that is,
$$g=dr^2+f(r)\sigma.$$
Notice that $S^n(H)$ has metric
$$g_H=dr^2+\sn_H^2(r)\sigma.$$
Plug this into $\vol(B_p(\rho))=\vol(B^H(\rho))$, we have
$$\int_0^\rho\int_{S^{n-1}}\sqrt{f(r)\det(g_{ij})}d\vol_{S^{n-1}}dr=\int_0^\rho\int_{S^{n-1}}\sqrt{\sn_H^2(r)\det(g_{ij})}d\vol_{S^{n-1}}dr.$$
Notice that
$$\int_{S^{n-1}}\sqrt{\det(g_{ij})}d\vol_{S^{n-1}}=\omega_{n-1},$$
we thus obtain
$$\int_0^\rho\sqrt{f(r)}dr=\int_0^\rho\sn_H(r)dr$$
holds for all $0<\rho<R$.
Therefore $f(r)=\sn_H^2(r)$ on $(0,R]$, and $B_p(R)$ is isometric to $B^H(R)$.
\advance\propcount by 1
\qed
Using~\ref{prop:volume rigidity}, we can derive volume rigidity for closed and open manifolds.
\medskip
\proclaim Theorem~\propnumber~{\rm(Volume rigidity for closed manifolds)}.
Let $(M,g)$ be a Riemannian manifold with $\Ric\geq n-1$.
If $\vol M=\vol S^{n-1}$, then $M$ is isometric to $S^n$.\par
\noindent{\it Proof.}\/ By Bonnet--Myers' theorem, we have $\diam M\leq\pi$, hence
$$\vol(B_p(\pi))=\vol M=\vol S^n=\vol(B^1(\pi)).$$
Then apply~\ref{prop:volume rigidity}.
\advance\propcount by 1
\qed
\proclaim Theorem~\propnumber~{\rm(Volume rigidity for open manifolds)}.
Let $(M,g)$ be a Riemannian manifold with $\Ric\geq 0$.
If
$$\lim_{\rho\to+\infty}{n\vol(B_p(\rho))\over\omega_n\rho^n}=1,$$
then $M$ is isometric to ${\bb R}^n$.\par
\noindent{\it Proof.}\/ The function
$$\rho\mapsto{n\vol(B_p(\rho))\over\omega_n\rho^n}={\vol(B_p(\rho))\over\vol(B^0(\rho))}$$
is nonincreasing and has limit $1$, hence
$$\vol(B_p(\rho))\geq\vol(B^0(\rho)).$$
By volume comparison inequality, the equality must hold, then apply~\ref{prop:volume rigidity}~to all $\rho>0$.
\advance\propcount by 1
\qed
\advance\subsectcount by 1
\noindent{\bf\the\sectcount.\the\subsectcount.~Diameter Rigidity.}
Our result on diameter rigidity for closed manifolds is Shiu-Yuen Cheng's maximal diameter theorem.
\medskip
\proclaim Theorem~\propnumber~{\rm(Maximal diameter)}.
Let $(M,g)$ be a Riemannian manifold with $\Ric\geq n-1$ and $\diam M=\pi$, then $M$ is isometric to $S^n$.\par
\noindent{\it Proof.}\/ First notice that $M$ is compact by Bonnet--Myers' theorem, so we can choose $p,q\in M$ such that $|pq|=\pi$.
Consider $B_p(\rho)$ and $B_q(\pi-\rho)$, by triangle inequality, we have $B_p(\rho)\cap B_q(\pi-\rho)=\emptyset$.
Hence by volume comparison inequality, we have
$$\eqalign{
    \vol M&\geq\vol(B_p(\rho))+\vol(B_q(\pi-\rho))\cr
    &={\vol(B_p(\rho))\over\vol(B^1(\rho))}\cdot\vol(B^1(\rho))+{\vol(B_q(\pi-\rho))\over\vol(B^1(\pi-\rho))}\cdot\vol(B^1(\pi-\rho))\cr
    &\geq{\vol(B_p(\pi))\over\vol(B^1(\pi))}\cdot\vol(B^1(\rho))+{\vol(B_q(\pi))\over\vol(B^1(\pi))}\cdot\vol(B^1(\pi-\rho))\cr
    &={\vol M\over\vol S^n}\cdot(\vol(B^1(\rho))+\vol(B^1(\pi-\rho)))\cr
    &={\vol M\over\vol S^n}\cdot\vol S^n\cr
    &=\vol M.
}$$
Therefore the equality of volume comparison inequality holds for $0<\rho\leq\pi$ on $M$, applying~\ref{prop:volume rigidity}~we obtain $B_p(\rho)$ is isometric to $B^1(\rho)$.
In particular, $M=B_p(\pi)$ is isometric to $B^1(\pi)=S^n$.
\advance\propcount by 1
\qed
We need to explain what does ``diameter rigidity'' mean for an open manifold.
Being of infinite diameter is not enough, there should be two points at infinity.
That is, there should be a geodesic joining two points at infinity, namely, a {\it line}\/ exists.
\medskip
\noindent{\bf Definition~\propnumber.} Let $\gamma:{\bb R}\to M$ be a geodesic, if for any $s,t\in{\bb R}$ the equality $|\gamma(s)\gamma(t)|=|s-t|$ holds, then $\gamma$ is called a {\it line}.
\advance\propcount by 1
\medskip
We expect rigidity result happens when a line exists. This leads to Cheeger--Gromoll's splitting theorem.
\medskip
\proclaim Theorem~\propnumber~{\rm(Splitting)}.
Let $(M,g)$ be a simply-connected Riemannian manifold with $\Ric\geq 0$, and a line exists on $M$.
Then $M$ is isometric to ${\bb R}\times N$ with product metric.\par
\advance\propcount by 1
Before proving this theorem, we need to input several big theorems.
\proclaim Theorem~\propnumber~{\rm(Strong maximum principle)}.
Let $(M,g)$ be a Riemannian manifold, $f$ be a Lipschitz function on $M$.
If $f$ satisfies $\Delta f\leq 0$ in the distribution sense and has a global minimum, then $f$ is constant on $M$.\par
\advance\propcount by 1
\proclaim Theorem~\propnumber~{\rm(Weyl's lemma)}.
Let $(M,g)$ be a Riemannian manifold, $f$ be a Lipschitz function.
If $\Delta f=0$ in the distribution sense, then $f$ is smooth.\par
\advance\propcount by 1
\proclaim Theorem~\propnumber~{\rm(De Rham decomposition)}.
Let $(M,g)$ be a simply-connected Riemannian manifold.
Let $\cal F$ be a distribution that is parallel with respective to the Levi--Civita connection, ${\cal F}^\perp$ be its orthogonal complement.
For any $p\in M$, let $N,N^\perp$ be the maximal integral submanifold of ${\cal F},{\cal F}^\perp$ at $p$.
Equip $N,N^\perp$ with induced metric from $M$, then $M$ is isometric to $N\times N^\perp$ with product metric.\par
\advance\propcount by 1
We will not prove these three theorems here.
\medskip
\noindent{\it Proof of splitting theorem.}\/ The proof uses {\it Busemann function}.
We first define and discuss some properties of Busemann function.

Fix $p\in M$, let $\gamma$ be a line on $M$.
For $t\in{\bb R}$, define $b^+_t(p)=|p\gamma(t)|-t$.
Then by triangle inequality,, if $t_1>t_2$, we have
$$b^+_{t_1}(p)-b^+_{t_2}=|p\gamma(t_1)|-|p(\gamma(t_2)|-(t_1-t_2)\leq 0,$$
hence $b^+_t(p)$ is nonincreasing with respective to $t$.
Moreover, by triangle inequality, we have $b^+_t(p)\geq-|p\gamma(0)|$, hence $b^+_t(p)$ is bounded below.
Thus we can take limit
$$B^+(p):=\lim_{t\to+\infty}b^+_t(p).$$
Similarly we define $b^-_t(p)=|p\gamma(-t)|-t$, we have the limit
$$B^-(p):=\lim_{t\to+\infty}b^-_t(p)$$
is well-defined.

Now we discuss the properties of Busemann functions.
First we notice that
$$B^+(p)+B^-(p)=|p\gamma(t)|+|p\gamma(-t)|-2t\geq 0.$$
Then by Laplacian comparison inequality, we have for any test function $\varphi\in C^\infty_0(M)$, $\varphi\geq 0$
$$\int_Mb^+_t\Delta\varphi\leq\int_M{n-1\over b^+_t+t}\cdot\varphi.$$
Let $t\to+\infty$, we obtain (we may use Lebesgue's dominated convergence theorem)
$$\int_MB^+\Delta\varphi\leq 0.$$
Similarly we have
$$\int_MB^-\Delta\varphi\leq 0.$$
Thus $\Delta(B^++B^-)\leq 0$ in the distribution sense.
Moreover, clearly $B^++B^-=0$ on $\gamma$, hence $B^++B^-$ satisfies the assumption of strong maximum principle, we have $B^++B^-=0$.
In particular, we have $\Delta(B^++B^-)=0$ in the classical sense, hence $\Delta B^+=\Delta B^-=0$ in the distribution sense.
By Weyl's lemma, we have $B^+$ and $B^-$ are both smooth.
Moreover, since the gradient of distance function has modulus $1$ almost everywhere, we have $|\nabla B^+|=1$.
Using Bochner's formula, we have
$$0=|\Hess{B^+}|^2+\langle\nabla B^+.\nabla\Delta B^+\rangle+\Ric(\nabla B^+,\nabla B^+)\geq|\Hess{B^+}|^2,$$
hence $\Hess{B^+}=0$.

The rest part is easy geometry.
Since covariant differetiation is commutative with musical isomorphism, we have $\nabla(\nabla B^+)=(\Hess{B^+})^\sharp=0$.
Hence $\nabla B^+$ is parallel.
Choose a regular point $p$ of $B^+$, by de Rham's decomposition theorem, $M$ is isometric to the product of integral curve of $\nabla B^+$ through $p$ and the level set of $B^+$ at $p$.
Without loss of generality we can assume $p$ lies on $\gamma$, then since $\nabla B^+$ is parallel, the integral curve is the geodesic through $p$, i.e.\ the line.
\qed
